Approximation as Differentiation

### 2025-07-02 main/introduction.tex
Rough stab at Galois slicing.

Perera and collaborators introduced the idea of {\em Galois program slicing} as a particular conception of program slicing for provenance, described in several publications~\cite{perera12a,perera16d,ricciotti17}. Galois program slicing (hereafter simply {\emph \GPS}) forms the basis of the open source data visualisation tool \href{https://f.luid.org/}{Fluid} that allows interactive exploration of programmatically generated visualisations.

>>>>

Perera and collaborators introduced the idea of {\em Galois program slicing} as a particular conception of program slicing for provenance, described in several publications~\cite{perera12a,perera16d,ricciotti17}. Galois program slicing (hereafter simply {\emph \GPS}) forms the basis of the open source data visualisation tool \href{https://f.luid.org/}{Fluid}\cite{perera2025fluid} that allows interactive exploration of programmatically generated visualisations.

### 2025-07-02 main/related-work.tex
Rough stab at Galois slicing.

\cite{perera12a,perera13,perera16d,ricciotti17,perera22,bond25}

>>>>

Galois slicing was introduced in \cite{perera12a} as an operational approach to program slicing based on the
idea of computing slices via Galois connections between lattices of input and output approximations. A
connection to Berry's stable functions in relation to minimal slices for short-circuiting operations was
alluded to in \citeauthor{perera13}'s PhD thesis but not elaborated upon. Subsequent work extended the
approach to functional languages with assignment and exceptions~\cite{ricciotti17} and concurrent systems,
applying Galois slicing to the $\pi$-calculus~\cite{perera16d}. In this latter work, the analysis shifted from
functions to transition relations, considering individual transitions $P \longrightarrow Q$ between specific
configurations $P$ and $Q$ as analogous to the ``edge'' between points $x$ and their images $f(x)$ in $f$, and
building Galois connections between $\downset{P}$ and $\downset{Q}$. This required generalizing the semantics
to capture behavioural provenance in the presence of nondeterminism. The main difference with the Galois
slicing approach presented here is that the earlier work also computes \emph{program slices}, using
approximation lattices that represent partially erased programs.

More recent work explored the application of Galois slicing in interactive visualisations. \citet{perera22}
presented an approach where slicing operates over Boolean algebras rather than plain lattices. In this setting
every Galois connection $f \dashv g: A \to B$ has a conjugate $g^\circ \dashv f^\circ: B \to A$, where
$f^\circ$ denotes the De Morgan dual $\neg \comp f \comp \neg$~\cite{jonsson51}. The provenance analysis can
then be composed with its own conjugate to obtain a Galois connection which computes \emph{related outputs}
(e.g., selecting a region of a chart and observing the regions of other charts which have overlapping data
dependencies). \citet{bond25} revisited this approach using \emph{dynamic dependence graphs} as an
implementation technique that decouples the derivation of dependency information from the analyses that make
use of it, and observing that to compute the conjugate analysis one can just use the opposite graph.
